---
title: "Week2B Classification Metrics: Requirements"
format: html
editor: visual
---

## Introduction

For this assignment, I will use the [`penguin_predictions.csv`](https://raw.githubusercontent.com/acatlin/data/refs/heads/master/penguin_predictions.csv) dataset to understand how well a classification model predicts whether a penguin is female or male. First, I will look at how many penguins are in each class and calculate the null error rate. This will give me a simple baseline to compare with the model.

Next, I will use the .pred_female column to make predictions using three different thresholds: 0.2, 0.5, and 0.8. I will create a confusion matrix for each threshold and calculate accuracy, precision, recall, and F1 score. Then, I will compare the results to see how changing the threshold affects the model.

### Data Challenges

Understanding how the different thresholds change the predictions. I also need to correctly identify true positives, false positives, true negatives, and false negatives.

### Implementation
```{r}
url<-"C:/Users/tenlh/Downloads/penguin_predictions.csv"
penguin_prediction<-read.csv(file=url)

```

```{r}
table(penguin_prediction$sex)

sum(table(penguin_prediction$sex))
```

## Task 1: Null error rate

```{r}
null_error_rate<-table(penguin_prediction$sex)["female"]/sum(table(penguin_prediction$sex))
null_error_rate
```
Null error rate is 41%

```{r}
ggplot(penguin_prediction,aes(x=sex))+
  geom_bar()
```

Null error rate is the error rate a classification model gets when it always predicts the single most common class in a data set.
It is important to know null error rate when evaluating model because, it acts as a baseline or benchmark. A good machine learning model must score an error rate lower than this null rate

## Task 2:Confusion Matrices at Multiple Thresholds
### Threshold =0.2
If .pred_female>0.2 "female","male"

Compute True Positives (TP)
```{r}
table(penguin_prediction$sex, ifelse(penguin_prediction$.pred_female>=0.2,1,0))

```
True Positive (TP) = 37
True Negative (TN) = 48
False Positive (FP) = 6
False Negative (FN) = 2
### Threshold =0.5
```{r}
table(penguin_prediction$sex, ifelse(penguin_prediction$.pred_female>=0.5,1,0))
```
True Positive (TP) = 36
True Negative (TN) = 51
False Positive (FP) = 3
False Negative (FN) = 3

### Threshold =0.8
```{r}
table(penguin_prediction$sex, ifelse(penguin_prediction$.pred_female>=0.8,1,0))
```
True Positive (TP) = 36
True Negative (TN) = 52
False Positive (FP) = 2
False Negative (FN) = 3

## Task 3: Performance Matrics
### Threshold=0.2
Accuracy = (TP + TN) / total
```{r}
accuracy<-(37+48)/(37+48+6+2)
accuracy
```
Accuracy is 91.39%
Precision = TP / (TP + FP)
```{r}
precision <- 37 / (37 + 6)
precision
```
precision is 86%

Recall = TP / (TP + FN)
```{r}
recall<-37/(37+2)
recall

```
recall is 94.87%
F1 = harmonic mean of precision and recall

F1 <- 2 * (precision * recall) / (precision + recall)

```{r}
F1 <- 2 * (0.8604651 *0.9487179 ) / (0.8604651 + 0.9487179)
F1

```
F1 score is 90.2%
### Threshold=0.5
```{r}
accuracy<-(36+51)/(36+51+3+3)
accuracy
```
Accuracy = (TP + TN) / total

Accuracy is 93.54%

```{r}
precision <- 36 / (36 + 3)
precision
```
precision is 92.3%

Recall = TP / (TP + FN)
```{r}
recall<-36/(36+3)
recall
```
recall is 92.3%


F1 <- 2 * (precision * recall) / (precision + recall)

```{r}
F1 <- 2 * (0.9230769 *0.9230769 ) / (0.9230769 + 0.9230769)
F1
```
F1 score is 92.3%
### Threshold=0.8
Accuracy = (TP + TN) / total
```{r}
accuracy<-(36+52)/(36+52+2+3)
accuracy
```
Accuracy is 94.62%
Precision = TP / (TP + FP)
```{r}
precision <- 36/ (36 + 2)
precision
```
precision is 94.73%

Recall = TP / (TP + FN)
```{r}
recall<-36/(36+2)
recall

```
recall is 94.73%
F1 = harmonic mean of precision and recall

F1 <- 2 * (precision * recall) / (precision + recall)

```{r}
F1 <- 2 * (0.9473684 *0.9473684 ) / (0.9473684 + 0.9473684)
F1

```
F1 score is 94.73%

## Threshold Use Cases
0.2 threshold is preferable when we need to identify as many people who have COVID-19.
0.8 is preferable when it is more important to make sure accuracy is high.Health care provider is more confident that a person is likely to have Covid-19
