---
title: "Assignment:5B ELO Calculations"
author: "Tenzin Lhamo Thakuri"
date: "2026-10-01"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task

Chess ELO calculations

Based on difference in ratings between the chess players and each of their opponents in our Project 1 tournament, calculate each player’s expected score (e.g. 4.3) and the difference from their actual score (e.g 4.0). List the five players who most overperformed relative to their expected score, and the five players that most underperformed relative to their expected score.

## Approach

Step 1: Calculate the expected score of each player using the ELO formula mentioned here <https://www.youtube.com/watch?v=AsYfbmp0To0> Step 2: calculate difference of players expected score from their actual score, in my data "Total_Points". Step 3: Sort result data in descending order to find five most overperformed players relative to their actual score. step 4: Sort result data in ascending order to find five most underperformed players relative to their actual score.

## Anticipated Challenges

-Find expected score of all the palyers, sort them to analyse it. 

## Code Base

```{r}
library(tidyverse)

# 1. Read and clean text
text <- read_lines("https://raw.githubusercontent.com/lhamo07/Data-607-Assignment/refs/heads/Project1/tournamentinfo.txt")

tournamentInfo_df <- tibble(text = text) %>%
  filter(!str_detect(text, "^\\s*-+\\s*$")) %>%
  slice(-(1:2)) %>%
  mutate(person = ceiling(row_number() / 2)) %>%
  group_by(person) %>%
  summarise(text = paste(text, collapse = " "), .groups = "drop") %>%
  mutate(
    Pair_Num = as.numeric(str_extract(text, "^\\s*(\\d+)")),
    Player_name = str_trim(str_extract(text, "(?<=\\|\\s)[A-Z]+(?:[-\\s]+[A-Z]+)+(?=\\s*\\|)")),
    Total_Points = as.numeric(str_extract(text, "(?<=\\|)\\s*\\d+\\.\\d+")),
    Player_pre_Rating = as.numeric(str_extract(text, "(?<=R:)\\s*\\d+"))
  )

# 2. Extract raw round text blocks and parse opponent ID numbers
rounds_df <- tournamentInfo_df %>%
  mutate(rounds_text = str_extract(text, "(\\|\\s*[WLDXHP\\s]\\s*\\d*\\s*){7}(?=\\|)")) %>%
  separate(rounds_text, into = c("drop", "R1", "R2", "R3", "R4", "R5", "R6", "R7"), sep = "\\|") %>%
  mutate(across(R1:R7, ~ as.numeric(str_extract(.x, "\\d+")))) %>%
  select(Pair_Num, Player_name, Total_Points, Player_pre_Rating, R1:R7)

# 3. Create a clean lookup table for player ratings
rating_lookup <- rounds_df %>% 
  select(Pair_Num, Player_pre_Rating)

# 4. Map opponent IDs to their respective pre-tournament ratings
opponent_info_df <- rounds_df %>%
  pivot_longer(cols = R1:R7, names_to = "Round", values_to = "Opponent_Num") %>%
  left_join(rating_lookup, by = c("Opponent_Num" = "Pair_Num")) %>%
  rename(Opponent_Pre_Rating = Player_pre_Rating.y, Player_pre_Rating = Player_pre_Rating.x) %>%
  
  # Pivot back into a wide format showing Opponent ID and Opponent Rating side-by-side per round
  pivot_wider(
    names_from = Round, 
    values_from = c(Opponent_Num, Opponent_Pre_Rating),
    names_glue = "{Round}_{.value}"
  ) %>%
  # Reorder columns chronologically (R1 info, R2 info, etc.)
  select(
    Pair_Num, Player_name, Total_Points, Player_pre_Rating,
     R1_Opponent_Pre_Rating,
   R2_Opponent_Pre_Rating,
   R3_Opponent_Pre_Rating,
   R4_Opponent_Pre_Rating,
     R5_Opponent_Pre_Rating,
     R6_Opponent_Pre_Rating,
     R7_Opponent_Pre_Rating
  )

opponent_info_df


```

### Calculate ELO

The standard Elo expected score formula is $$E_A = \frac{1}{1 + 10^{(R_B - R_A) / 400}}$$

Where: \* $R_A$ is Player A's current rating. \* $R_B$ is Player B's current rating.

```{r}
elo_df<-opponent_info_df%>%
# Drops any row where Rating_A or Rating_B is missing

drop_na(R1_Opponent_Pre_Rating, R2_Opponent_Pre_Rating, R3_Opponent_Pre_Rating,
          R4_Opponent_Pre_Rating, R5_Opponent_Pre_Rating, R6_Opponent_Pre_Rating, 
          R7_Opponent_Pre_Rating)%>%  mutate(ELO_R1=1/(1+10^((R1_Opponent_Pre_Rating-Player_pre_Rating)/400)),
         ELO_R2=1/(1+10^((R2_Opponent_Pre_Rating-Player_pre_Rating)/400)),
         ELO_R3=1/(1+10^((R3_Opponent_Pre_Rating-Player_pre_Rating)/400)),
         ELO_R4=1/(1+10^((R4_Opponent_Pre_Rating-Player_pre_Rating)/400)),
         ELO_R5=1/(1+10^((R5_Opponent_Pre_Rating-Player_pre_Rating)/400)),
         ELO_R6=1/(1+10^((R6_Opponent_Pre_Rating-Player_pre_Rating)/400)),
         ELO_R7=1/(1+10^((R7_Opponent_Pre_Rating-Player_pre_Rating)/400)),
         ELO_total=ELO_R1+ELO_R2+ELO_R3+ELO_R4+ELO_R5+ELO_R6+ELO_R7
         
         
         )
elo_df
```

### Calculate difference from actual score

```{r}
final_df<-elo_df%>%
  mutate(Score_minus_Expected=Total_Points-ELO_total)%>%
  select(Pair_Num,Player_name,Total_Points,Player_pre_Rating,ELO_total,Score_minus_Expected)

final_df
```

### Calculate five most overperformed relative to their expected score

```{r}
overperform_df<-final_df%>%
  arrange(desc(Score_minus_Expected))
head(overperform_df,n=5)
```

### Calculate five most underperformed relative to their expected score

```{r}
underperform_df<-final_df%>%
  arrange(Score_minus_Expected)
head(underperform_df,n=5)
```

## Conclusion

Based on the overperformed relative to their expected score data, Aditya Bajaj scored 6.0 total points even though his predicted score was only 1.95 points. This means Aditya performed much better than expected. Similarly, Zachary James Houghton and Anvit Rao also performed significantly better than the model predicted. Based on the underperformed relative to their expected score data, Loren Schwiebert scored only 3.5 total points even though his predicted score was 6.28 points. This means he performed much worse than expected. Similarly, George Avery Jones, Jared Ge, Rishi Shetty, and Joshua David Lee also scored fewer points than the model predicted.

## Citation

[Large language model]. google.com. Accessed October 4, 2026. <https://share.google/aimode/j7NEiRnQ1M8CZXzpC>
