---
title: 'Assignment 5A: Airline Delays'
author: "Tenzin Lhamo Thakuri"
date: "2026-09-30"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task

Submitted data file with at least some count analysis (+45)

Recreated File in Same Format as given, including missing data where
there were empty cells in source data. (+5)

Placed file in Internet-accessible location, such as a
publicly-accessible GitHub repo. Also OK (if less scalable) to
generated and populated data frame from code. (+5)

Provided code to populate missing data. (+5)

Transformed data from wide to long format. (+10)

Compared percentage (not just counts) of either delays or arrival rates
for two airlines overall. [Either with a chart, or a table; both would
be exemplary]. Include text summarizes findings from this comparison.
(+10)

Compared percentage (not just counts) of either delays or arrival rates
for two airlines across five cities. [Either with a chart, or a table;
both would be exemplary]. Include text summarizes findings from this
comparison. (+10)

Describe discrepancy between comparing two airlines’ flight performances
city-by-city and overall. (+5)

Assignment Tidying and Transforming Data

Explain discrepancy between comparing two airlines’ flight performances
city-by-city and overall.(+5)

The chart above describes arrival delays for two airlines across five
destinations. Your task is to: (1) Create a .CSV file (or optionally, a
MySQL database!) that includes all of the information above. You’re
encouraged to use a “wide” structure similar to how the information
appears above, so that you can practice tidying and transformations as
described below. (2) Read the information from your .CSV file into R,
and use tidyr and dplyr as needed to tidy and transform your data. (3)
Perform analysis to compare the arrival delays for the two airlines.

## Approach

Step 1: Create database and table to insert given two airlines delay
data (wide structure data) in postgreSQL. Step 2: Connect R to
PostgreSQL Step 3: Read a SQL table into R Step 4: Save the SQL table as
CSV Step 5: Read wide structure data from github into R. Step 6:
Transform wide data to long structure data. Step 7: Compare two airlines
arrival delays across five cities using plot percentage

## Anticipated Challenges

converting wide structure data to long structure. Making comparison
between two airlines across all cities.

## Code Base
### Load necessary libraries
```{r}
library(DBI)
library(RPostgres)
library(tidyverse)
```
### Create Database, Table, and Insert Data

The following SQL code creates the airlines_arrival_delay table, inserts the data provided in the assignment, and displays all the records in the table.

CREATE TABLE airlines_arrival_delay("airline" VARCHAR(50), status VARCHAR(50), los_angeles Int,phoenix Int, san_diego Int,san_francisco Int,seattle Int);
INSERT INTO airlines_arrival_delay(airline,status,los_angeles,phoenix,san_diego,san_francisco,seattle)VALUES
('ALASKA','on time',497,221,212,503,1841),
(NULL,'delayed',62,12,20,102,305),
(NULL, NULL, NULL, NULL, NULL, NULL, NULL),
('AMWEST','on time',694,4840,383,320,201),
(NULL,'delayed',117,415,65,129,61);
SELECT * FROM airlines_arrival_delay;

### Connect R to PostgreSQL
```{r}
con <- dbConnect(
  RPostgres::Postgres(),
  dbname = "Airlines_delay",
  host = "localhost",
  port = 5432,
  user = "postgres",
  password = Sys.getenv("DB_PASSWORD")
)
```

```{r}
dbListTables(con)
```

```{r}
airlines_df <- dbReadTable(con, "airlines_arrival_delay")
head(airlines_df)
```
```{r}
#set first and second columns name empty.
names(airlines_df)[1:2] <- c(" ", " ")
write.csv(airlines_df, file="airlines_arrival_delay.csv", row.names = FALSE)
```
### Load airline arrival delay dataset
```{r}
url<-"https://raw.githubusercontent.com/lhamo07/Data-607-Assignment/refs/heads/main/Assignment-5A/airlines_arrival_delay.csv"
airlines_arrival_delay_df<-read.csv(file=url)
airlines_arrival_delay_df
```
Rename the first two un-labelled columns as "airline" and "status"

```{r}
colnames(airlines_arrival_delay_df)[1:2] <- c("airline", "status")
airlines_arrival_delay_df
```

```{r}
airline_long_structure<-airlines_arrival_delay_df%>%
  fill(airline)%>%
  pivot_longer(cols=!(airline:status),
               names_to="city",
               values_to="flight_count",
                   values_drop_na = TRUE

               )
airline_long_structure
```
First, I used fill(airline) to fill in the missing airline names with the corresponding airline name above. Then, I used pivot_longer() to convert the data from a wide structure to a long structure. In cols = !(airline:status), all columns except the columns from airline through status are selected for transformation. The selected column names are stored in the new city column using names_to = "city", and their corresponding values are stored in the flight_count column using values_to = "flight_count". Finally, values_drop_na = TRUE removes rows where the flight count value is missing NA
### Percentage of delays rates for ALASKA airlines
```{r}
alaska_summary<-airline_long_structure%>%
filter(airline=="ALASKA")%>%
  summarize(
    airline="ALASKA",
    total_delayed=sum(flight_count[status=='delayed']),
    total_on_time = sum(flight_count[status == "on time"]),
    grand_total   = total_delayed + total_on_time,
    delay_percent = round((total_delayed / grand_total) * 100,digits=2)
    )
alaska_summary


```
###  Percentage of delays rates for AMWEST airlines

```{r}
amwest_summary<-airline_long_structure%>%
filter(airline=="AMWEST")%>%
  summarize(
    airline="AMWEST",
    total_delayed=sum(flight_count[status=='delayed']),
    total_on_time = sum(flight_count[status == "on time"]),
    grand_total   = total_delayed + total_on_time,
    delay_percent = round((total_delayed / grand_total) * 100,digits=2)
    )
amwest_summary


```

```{r}
percentage_summary<-bind_rows(alaska_summary,amwest_summary)
percentage_summary
```
### Compared percentage of delays  rates for two airlines overall.

```{r}
ggplot(percentage_summary,aes(x=airline,y=delay_percent,fill=airline))+
  geom_col(width=0.5)+
   geom_text(
    aes(label = paste0(delay_percent, "%")),
    vjust = -0.5)+
    labs(
      title = "Overall Arrival Delay Percentage by Airline",
    x = "Airline",
    y = "Overall Delay Percentage (%)"
    )+
    theme_minimal()
```

From the above plot, I found that the overall delay percentage for ALASKA Airline is 13.27%, while AMWEST Airline is 10.83%. Based on this comparison, I would say that AMWEST Airlines appears to be more reliable for passengers who are concerned about flight delays.

### Compared percentage of delays rates for two airlines across five cities.
```{r}
city_summary<-airline_long_structure%>%
group_by(airline,city)%>%
    mutate(city_total=sum(flight_count),
           delay_percentage=round((flight_count/city_total)*100,digits=2))%>%
  

  filter(status=="delayed")
  
 
city_summary
```

```{r}
ggplot(city_summary,aes(x=city,y=delay_percentage,fill=airline))+
  geom_col(width=0.5,position="dodge")+
  geom_text(
    aes(label = paste0(delay_percentage, "%")),
    vjust = -0.5)+
    labs(
      title = "Overall Arrival Delay Percentage by Airline",
    x = "Airline",
    y = "Overall Delay Percentage (%)"
    )+

    theme_minimal()
```

From the above plot, I can conclude that Phoenix has the lowest flight delay percentage for both airlines. San Francisco has the highest delay percentage for AMWEST Airlines. The remaining cities have relatively similar flight delay percentages for both airlines.

## Conclusion
Although the overall percentage of delays for ALASKA Airlines is higher than AMWEST Airlines, the city level comparison shows that AMWEST has a higher delay percentage in some cities. Therefore, if I were planning a trip, I would compare the delay percentages by city rather than relying only on the overall delay percentage.

### citation
https://chatgpt.com/share/6abc7bac-d99c-83e9-a515-e584ba7ef430
https://chatgpt.com/share/6ac2d87e-7cb8-83ea-899e-9e8371c0fbb0
OpenAI. (2026). ChatGPT [Large language model]. https://chatgpt.com. Accessed October 4, 2026.